---
title: "Nonlinear Models for the S&P 500"
subtitle: "TAR, Markov Switching, Nonlinearity Tests, Parametric Bootstrap and Forecast Evaluation"
author: "FI362 --- Advanced Financial Econometrics"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    theme: readable
  pdf_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 4.2, cache = FALSE)
options(digits = 5)
```

# What this notebook does, and what it refuses to do

This notebook implements, on real S&P 500 data, the pipeline in the accompanying
lecture: fit a linear benchmark, work out **where** the nonlinearity lives
(mean or variance), fit a SETAR and a Markov switching model, test linearity
**by parametric bootstrap** rather than against a $\chi^2$ table, and then
evaluate forecasts with tools that can actually see what a nonlinear model buys
you.

Three commitments, stated up front because they are the places where published
work most often goes wrong:

1. **Every core routine is written in base R.** Contributed packages (`tsDyn`,
   `MSwM`) appear only as cross-checks. If you cannot see the arithmetic, you
   cannot defend the result.
2. **The bootstrap re-runs the entire search inside every replication.** Fixing
   $\hat c$ inside the loop turns $\sup_c \mathrm{LR}(c)$ back into a pointwise
   $\chi^2$ statistic and makes the test over-reject.
3. **The Diebold--Mariano test is never applied to nested models.** An AR($p$) is
   nested in a SETAR. Use Clark--West.

```{r config}
CFG <- list(
  ticker      = "^GSPC",
  start_date  = "2000-01-01",
  end_date    = format(Sys.Date()),
  p_max       = 5,      # max AR order
  d_max       = 3,      # max threshold delay
  trim        = 0.15,   # threshold grid trimmed to [0.15, 0.85] quantiles
  B_boot      = 499,    # bootstrap replications (999 or 1999 for final results)
  burn_in     = 300,
  window      = 1000,   # rolling estimation window
  max_origins = 1000,   # cap on OOS origins (run time)
  refit_every = 20,
  seed        = 20260710
)
set.seed(CFG$seed)

have <- function(pkg) requireNamespace(pkg, quietly = TRUE)
opt <- list(quantmod = have("quantmod"), tseries = have("tseries"),
            tsDyn = have("tsDyn"), MSwM = have("MSwM"))
opt
```

---

# Data: from prices to log returns

We use $r_t = 100 \times \Delta \log P_t$, in **percent**.

Why percent, and not decimals? Purely numerical. In decimals the conditional
variance of a daily return is around $10^{-4}$; the Markov switching likelihood
then has parameters separated by four orders of magnitude and `optim()` fails to
converge. Scaling by 100 is not cosmetic --- it is the difference between a
model that estimates and one that does not.

```{r data}
get_sp500 <- function(cfg = CFG) {
  if (opt$quantmod) {
    out <- try({
      px <- quantmod::getSymbols(cfg$ticker, src = "yahoo",
                                 from = cfg$start_date, to = cfg$end_date,
                                 auto.assign = FALSE)
      px <- stats::na.omit(quantmod::Cl(px))
      list(dates = as.Date(zoo::index(px))[-1],
           r = as.numeric(100 * diff(log(as.numeric(px)))))
    }, silent = TRUE)
    if (!inherits(out, "try-error")) return(out)
  }
  message("Using a simulated fallback series (no data connection).")
  n <- 6000; P <- matrix(c(0.99, 0.01, 0.06, 0.94), 2, 2, byrow = TRUE)
  s <- integer(n); s[1] <- 1L
  for (t in 2:n) s[t] <- sample.int(2, 1, prob = P[s[t - 1], ])
  mu <- c(0.05, -0.10); sg <- c(0.70, 2.30)
  list(dates = seq.Date(as.Date(cfg$start_date), by = "day", length.out = n),
       r = mu[s] + sg[s] * rnorm(n))
}

dat <- get_sp500(); r <- dat$r; n <- length(r)
c(T = n, first = as.character(min(dat$dates)), last = as.character(max(dat$dates)))
```

## Stylised facts

```{r stylised}
round(c(
  mean     = mean(r),
  sd       = sd(r),
  skewness = mean((r - mean(r))^3) / sd(r)^3,
  kurtosis = mean((r - mean(r))^4) / sd(r)^4,
  acf1_r   = as.numeric(acf(r,   1, plot = FALSE)$acf[2]),
  acf1_r2  = as.numeric(acf(r^2, 1, plot = FALSE)$acf[2])
), 4)
```

The number to look at is the gap between `acf1_r` (near zero) and `acf1_r2`
(large). Returns are close to a martingale difference **in the mean** but
strongly dependent **in the square**. That single fact predicts almost everything
that follows: the mean-nonlinearity tests will be borderline, the variance tests
will reject overwhelmingly, and the MSFE gains from nonlinear mean models will
be negligible.

```{r plot-returns, fig.height=3.2}
plot(dat$dates, r, type = "l", col = "steelblue",
     xlab = "", ylab = "r (%)", main = "S&P 500 daily log returns")
abline(h = 0, col = "grey60")
```

---

# The linear benchmark

`make_lags(y, mlag)` discards the first `mlag` observations. The argument exists
for one reason: an AR($p$) and a SETAR with delay $d>p$ must be estimated on the
**same effective sample**, or the likelihood ratio between them is nonsense.

```{r ar}
make_lags <- function(y, mlag) {
  Z <- embed(y, mlag + 1)              # col 1 = y_t, col j+1 = y_{t-j}
  list(y = Z[, 1], X = cbind(1, Z[, -1, drop = FALSE]))
}

fit_ar <- function(y, p, mlag = p) {
  L <- make_lags(y, mlag)
  X <- cbind(1, L$X[, 2:(p + 1), drop = FALSE])
  b <- qr.solve(X, L$y)
  a <- as.numeric(L$y - X %*% b); neff <- length(a)
  list(p = p, mlag = mlag, coef = as.numeric(b), resid = a,
       ssr = sum(a^2), sigma2 = sum(a^2)/neff, n = neff,
       aic = neff * log(sum(a^2)/neff) + 2 * (p + 1))
}

aics <- sapply(1:CFG$p_max, function(p) fit_ar(r, p)$aic)
p_hat <- which.min(aics); ar0 <- fit_ar(r, p_hat)
cat("selected p =", p_hat, "\n"); round(ar0$coef, 5)
```

---

# Where is the nonlinearity? Mean or variance?

This ordering is not pedantry. BDS, Tsay's $F$ and the sup-LR test all reject
against a *general* alternative. If you have not first removed conditional
heteroskedasticity, a GARCH effect will present itself as a threshold in the
mean and you will fit --- and publish --- the wrong model.

## Engle's ARCH-LM and McLeod--Li

```{r arch}
arch_lm <- function(a, q = 12) {
  Z <- embed(a^2, q + 1); y <- Z[, 1]; X <- cbind(1, Z[, -1, drop = FALSE])
  fit <- lm.fit(X, y)
  r2 <- 1 - sum(fit$residuals^2) / sum((y - mean(y))^2)
  stat <- length(y) * r2
  c(stat = stat, df = q, p = pchisq(stat, q, lower.tail = FALSE))
}
round(arch_lm(ar0$resid), 4)
bt <- Box.test(ar0$resid^2, lag = 12, type = "Ljung-Box")   # McLeod-Li
c(stat = unname(bt$statistic), p = unname(bt$p.value))
```

Both reject at any conventional level. That is *variance* nonlinearity.

## Tsay's (1989) threshold $F$ test

The trick: sort the AR design matrix by the switching variable $y_{t-d}$, run
**recursive** least squares on the sorted rows, and test whether the predictive
residuals are orthogonal to the regressors. If the AR is stable across regimes
they are. The threshold $c$ never has to be estimated --- which is exactly why
this statistic has a standard $F$ limit while $\sup_c \mathrm{LR}(c)$ does not.

```{r tsay}
tsay_F <- function(y, p, d, start_frac = 0.10) {
  L <- make_lags(y, max(p, d))
  yy <- L$y
  X  <- cbind(1, L$X[, 2:(p + 1), drop = FALSE])
  z  <- L$X[, d + 1]                       # switching variable y_{t-d}
  o  <- order(z); yy <- yy[o]; X <- X[o, , drop = FALSE]
  N  <- length(yy); k <- ncol(X); m0 <- max(floor(start_frac * N), k + 2)

  ehat <- numeric(N - m0)
  Xi <- X[1:m0, , drop = FALSE]; yi <- yy[1:m0]
  Vi <- solve(crossprod(Xi)); b <- Vi %*% crossprod(Xi, yi)
  for (i in seq_len(N - m0)) {             # recursive least squares
    xn <- X[m0 + i, , drop = FALSE]
    fi <- as.numeric(1 + xn %*% Vi %*% t(xn))
    ehat[i] <- as.numeric(yy[m0 + i] - xn %*% b) / sqrt(fi)
    Kg <- Vi %*% t(xn) / fi                # Sherman-Morrison update
    b  <- b + Kg * as.numeric(yy[m0 + i] - xn %*% b)
    Vi <- Vi - Kg %*% (xn %*% Vi)
  }
  Xe <- X[(m0 + 1):N, , drop = FALSE]
  ssr1 <- sum(lm.fit(Xe, ehat)$residuals^2)
  ssr0 <- sum(ehat^2)
  Ne <- length(ehat)
  Fs <- ((ssr0 - ssr1)/k) / (ssr1/(Ne - k))
  c(F = Fs, p = pf(Fs, k, Ne - k, lower.tail = FALSE))
}

t(sapply(1:CFG$d_max, function(d) c(d = d, round(tsay_F(r, p_hat, d), 4))))
```

## BDS: before and after filtering the volatility

```{r bds, eval=opt$tseries}
tseries::bds.test(ar0$resid, m = 3, eps = sd(ar0$resid))
```

If you have `rugarch` or `fGarch` installed, refit a GARCH(1,1) and re-run BDS
on the **standardised** residuals $\hat a_t / \hat h_t$. The typical outcome on
S&P 500 data is a large drop in the BDS statistic. Interpretation: most of the
"nonlinearity" was volatility clustering. A threshold in the mean is a *residual*
claim that must survive this filtering.

---

# SETAR by conditional least squares

If $c$ were known, a SETAR would be two OLS regressions on two subsamples. It is
not known, so we **profile it out**: compute the concentrated SSR on a trimmed
grid of candidate thresholds and take the minimiser. Trimming at 15% in each tail
guarantees each regime retains enough observations.

$\hat c$ converges at rate $T$ (super-consistent), the slopes at rate
$\sqrt{T}$. Practical consequence: report slope standard errors *as if $c$ were
known*. Never the reverse.

```{r setar-fns}
setar_ssr <- function(y, p, d, c_val, mlag = max(p, d)) {
  L <- make_lags(y, mlag); yy <- L$y
  X <- cbind(1, L$X[, 2:(p + 1), drop = FALSE]); z <- L$X[, d + 1]
  i1 <- z <= c_val
  if (sum(i1) < ncol(X) + 5 || sum(!i1) < ncol(X) + 5) return(NA_real_)
  sum(lm.fit(X[i1, , drop = FALSE], yy[i1])$residuals^2) +
  sum(lm.fit(X[!i1, , drop = FALSE], yy[!i1])$residuals^2)
}

setar_grid <- function(y, p, d, trim = CFG$trim, ngrid = 200, mlag = max(p, d)) {
  L <- make_lags(y, mlag); z <- L$X[, d + 1]
  g <- unique(quantile(z[z >= quantile(z, trim) & z <= quantile(z, 1 - trim)],
                       probs = seq(0, 1, length.out = ngrid)))
  s <- vapply(g, function(cc) setar_ssr(y, p, d, cc, mlag), numeric(1))
  k <- which.min(s)
  list(c_hat = as.numeric(g[k]), ssr = as.numeric(s[k]),
       grid = as.numeric(g), ssr_path = as.numeric(s), n = length(L$y))
}

setar_fit <- function(y, p, d, c_val, mlag = max(p, d)) {
  L <- make_lags(y, mlag); yy <- L$y
  X <- cbind(1, L$X[, 2:(p + 1), drop = FALSE]); z <- L$X[, d + 1]
  i1 <- z <= c_val
  f1 <- lm.fit(X[i1, , drop = FALSE], yy[i1])
  f2 <- lm.fit(X[!i1, , drop = FALSE], yy[!i1])
  list(p = p, d = d, c = c_val, coef1 = f1$coefficients, coef2 = f2$coefficients,
       sigma1 = sqrt(mean(f1$residuals^2)), sigma2 = sqrt(mean(f2$residuals^2)),
       n1 = sum(i1), n2 = sum(!i1), n = length(yy),
       ssr = sum(f1$residuals^2) + sum(f2$residuals^2))
}
```

## Selecting $(p, d)$

All candidates share one effective sample (`MLAG`) so their AICs are comparable.

**A warning that belongs here and nowhere else:** this AIC is *not* a test of
linearity. The SETAR was chosen by minimising over $c$; the search itself
inflates the fit. Comparing $\mathrm{AIC}_{\text{SETAR}}$ to
$\mathrm{AIC}_{\text{AR}}$ and declaring victory is a selection effect. That is
what the bootstrap in the next section prices.

```{r setar-select}
MLAG <- max(CFG$p_max, CFG$d_max)
sel <- expand.grid(p = 1:CFG$p_max, d = 1:CFG$d_max)
sel$aic <- NA_real_; sel$c_hat <- NA_real_
for (i in seq_len(nrow(sel))) {
  g  <- setar_grid(r, sel$p[i], sel$d[i], mlag = MLAG)
  fi <- setar_fit(r, sel$p[i], sel$d[i], g$c_hat, mlag = MLAG)
  sel$aic[i]   <- fi$n * log(fi$ssr / fi$n) + 2 * (2 * (sel$p[i] + 1) + 2)
  sel$c_hat[i] <- g$c_hat
}
best <- sel[which.min(sel$aic), ]; best
setar_hat <- setar_fit(r, best$p, best$d, best$c_hat, mlag = MLAG)
unlist(setar_hat[c("c", "sigma1", "sigma2", "n1", "n2")])
```

Read `sigma1` against `sigma2`. With $\hat c$ near zero, $\hat\sigma_1 >
\hat\sigma_2$ says that days following a fall are more volatile than days
following a rise --- the leverage effect, appearing inside a *mean-equation*
model with no GARCH term anywhere.

```{r ssr-profile, fig.height=3.4}
gg <- setar_grid(r, best$p, best$d, mlag = MLAG)
plot(gg$grid, gg$ssr_path, type = "l", col = "steelblue", lwd = 2,
     xlab = "candidate threshold c", ylab = "concentrated SSR",
     main = "Profile SSR: the objective the grid search minimises")
abline(v = gg$c_hat, lty = 2, col = "firebrick")
```

The sharpness of the minimum is informative. A flat profile means $c$ is poorly
identified and any regime interpretation is fragile, whatever the $p$-value says.

## Cross-check with `tsDyn`

`tsDyn::setar()` indexes the delay from **zero**: `thDelay = d - 1`.

```{r tsdyn, eval=opt$tsDyn}
chk <- tsDyn::setar(r, m = best$p, thDelay = best$d - 1, nthresh = 1,
                    trim = CFG$trim)
c(tsDyn = round(tsDyn::getTh(chk), 4), ours = round(best$c_hat, 4))
```

---

# Markov switching: Hamilton filter and Kim smoother

$$
r_t = \mu_{s_t} + \phi\, r_{t-1} + \sigma_{s_t}\varepsilon_t, \qquad
P = \begin{pmatrix} p_{11} & 1-p_{11} \\ 1-p_{22} & p_{22}\end{pmatrix}.
$$

Intercept and variance switch; the AR slope is common. Two implementation points
that decide whether this converges:

* **Reparametrise.** Variances through `exp`, probabilities through `plogis`, so
  `optim()` is unconstrained. Forgetting this is the single most common cause of
  "the MS model would not estimate".
* **Break label switching.** $(\mu_1,\sigma_1,p_{11}) \leftrightarrow
  (\mu_2,\sigma_2,p_{22})$ gives an identical likelihood. Impose
  $\sigma_1 < \sigma_2$ *after* optimising, so state 1 is always "calm".

The filter is two lines. Predict, then apply Bayes. The normalising constant of
the Bayes step *is* the one-step predictive density, so the log-likelihood comes
out of the filter for free.

```{r ms-fns}
ms_unpack <- function(par) list(mu = par[1:2], phi = par[3], sig = exp(par[4:5]),
                                p11 = plogis(par[6]), p22 = plogis(par[7]))

ms_filter <- function(par, y) {
  th <- ms_unpack(par)
  P  <- matrix(c(th$p11, 1 - th$p11, 1 - th$p22, th$p22), 2, 2, byrow = TRUE)
  n <- length(y); den <- 2 - th$p11 - th$p22
  xi <- c((1 - th$p22)/den, (1 - th$p11)/den)          # ergodic start
  xp <- xf <- matrix(NA_real_, n, 2); llv <- numeric(n)
  for (t in 2:n) {
    pred <- as.vector(crossprod(P, xi))                # xi_{t|t-1} = P' xi_{t-1|t-1}
    eta  <- dnorm(y[t], th$mu + th$phi * y[t - 1], th$sig)
    num  <- pred * eta; f <- sum(num)
    if (!is.finite(f) || f <= 0) return(NULL)
    xp[t, ] <- pred; xf[t, ] <- num/f; llv[t] <- log(f); xi <- num/f
  }
  list(loglik = sum(llv[-1]), xi_pred = xp, xi_filt = xf, P = P, theta = th)
}
ms_negll <- function(par, y) { o <- ms_filter(par, y)
  if (is.null(o) || !is.finite(o$loglik)) 1e10 else -o$loglik }

# Kim (1994): xi_{t|T} = xi_{t|t} * [ P %*% (xi_{t+1|T} / xi_{t+1|t}) ]
# Note P, not P': we sum over the FUTURE state for a fixed current state.
ms_smooth <- function(flt) {
  xf <- flt$xi_filt; xp <- flt$xi_pred; P <- flt$P; n <- nrow(xf)
  xs <- matrix(NA_real_, n, 2); xs[n, ] <- xf[n, ]
  for (t in (n - 1):2) {
    xs[t, ] <- xf[t, ] * as.vector(P %*% (xs[t + 1, ] / xp[t + 1, ]))
    xs[t, ] <- xs[t, ] / sum(xs[t, ])
  }
  xs
}
```

```{r ms-fit}
start <- c(0.05, -0.10, 0, log(0.7), log(2.3), qlogis(0.98), qlogis(0.90))
fit_ms <- optim(start, ms_negll, y = r, method = "BFGS",
                control = list(maxit = 2000, reltol = 1e-10))
th <- ms_unpack(fit_ms$par)
if (th$sig[1] > th$sig[2]) {                       # enforce state 1 = calm
  fit_ms$par[c(1,2)] <- fit_ms$par[c(2,1)]
  fit_ms$par[c(4,5)] <- fit_ms$par[c(5,4)]
  fit_ms$par[c(6,7)] <- fit_ms$par[c(7,6)]
  th <- ms_unpack(fit_ms$par)
}
flt <- ms_filter(fit_ms$par, r); sm <- ms_smooth(flt)

den <- 2 - th$p11 - th$p22
data.frame(
  quantity = c("mu1","mu2","sigma1","sigma2","phi","p11","p22",
               "E[dur 1]","E[dur 2]","pi1","pi2","logLik"),
  value = round(c(th$mu, th$sig, th$phi, th$p11, th$p22,
                  1/(1 - th$p11), 1/(1 - th$p22),
                  (1 - th$p22)/den, (1 - th$p11)/den, flt$loglik), 4))
```

## Three sanity checks, in order

1. $p_{11}, p_{22} > 0.9$. Regimes should be *persistent states*, not dummies for
   outliers.
2. $1/(1-\hat p_{22}) > 5$ days. If the turbulent regime dies in two days it is
   picking up jumps, not volatility clustering. Remedy: allow GARCH within
   regimes.
3. $\pi_2$ roughly $0.1$--$0.3$. A turbulent regime occupying half the sample is
   not a crisis regime.

## Crisis dating

The smoothed probability of the high-variance regime is the model's own,
entirely data-driven, crisis chronology. Check it against 2008, March 2020 and
2022. If it does not line up, something is wrong --- with the model, or with the
data.

```{r ms-plot, fig.height=3.4}
plot(dat$dates[-1], sm[-1, 2], type = "l", col = "firebrick", ylim = c(0, 1),
     xlab = "", ylab = expression(Pr(s[t] == 2 ~ "|" ~ F[T])),
     main = "Smoothed probability of the turbulent regime")
abline(h = 0.5, lty = 2)
```

```{r ms-dating}
crisis <- dat$dates[which(sm[, 2] > 0.5)]
round(mean(sm[-1, 2] > 0.5), 3)          # share of days in the turbulent state
head(sort(table(format(crisis, "%Y")), decreasing = TRUE), 6)
```

---

# Testing linearity: the Davies problem and the parametric bootstrap

Under $H_0$ (AR($p$)) the likelihood is **exactly flat in $c$**. The threshold is
a nuisance parameter present only under the alternative. Consequently
$\mathrm{LR}(c)$ is a *stochastic process* indexed by $c$, and
$\sup_c \mathrm{LR}(c)$ converges to the supremum of a squared Gaussian process
whose covariance kernel depends on the DGP and on the grid. It is **not**
$\chi^2$, and because a supremum can only increase a statistic, the true null
distribution lies to the *right* of the $\chi^2$ you were about to use.

```{r suplr}
sup_LR <- function(y, p, d, trim = CFG$trim, mlag = max(p, d)) {
  ar_f <- fit_ar(y, p, mlag = mlag)                 # same effective sample
  g    <- setar_grid(y, p, d, trim, mlag = mlag)    # ... as the SETAR
  g$n * (log(ar_f$ssr) - log(g$ssr))
}
```

## Simulating the null

The innovation scheme is not a detail. Daily S&P 500 returns have strong ARCH.
An i.i.d. residual bootstrap reshuffles $\hat a_t$ and **destroys volatility
clustering**: the bootstrap samples are too calm, extreme values of $y_{t-d}$
that drive large $\sup_c \mathrm{LR}$ are too rare, the simulated null is too
tight, and the test **over-rejects**. You will find a threshold where there is
only GARCH.

The wild bootstrap $a_t^* = \hat a_t \eta_t$ preserves the conditional variance
pattern and is the safe default. Better still: bootstrap under an AR--GARCH null.

```{r boot}
sim_ar_null <- function(arfit, T_out, burn = CFG$burn_in,
                        scheme = c("wild","residual","parametric")) {
  scheme <- match.arg(scheme); p <- arfit$p; b <- arfit$coef
  N <- T_out + burn + p
  e <- switch(scheme,
    parametric = rnorm(N, 0, sqrt(arfit$sigma2)),
    residual   = sample(arfit$resid, N, replace = TRUE),
    wild       = sample(arfit$resid, N, replace = TRUE) * rnorm(N))
  y <- numeric(N); dn <- 1 - sum(b[-1])
  y[1:p] <- if (abs(dn) > 1e-6) b[1]/dn else 0
  for (t in (p + 1):N) y[t] <- b[1] + sum(b[-1] * y[(t - 1):(t - p)]) + e[t]
  y[(burn + p + 1):N]
}

boot_supLR <- function(y, p, d, B = CFG$B_boot, scheme = "wild",
                       mlag = max(p, d)) {
  obs <- sup_LR(y, p, d, mlag = mlag)
  arf <- fit_ar(y, p, mlag = mlag); Tn <- length(y)
  stars <- vapply(seq_len(B), function(b) {
    ys <- sim_ar_null(arf, Tn, scheme = scheme)
    sup_LR(ys, p, d, mlag = mlag)     # <-- FULL procedure, grid search included
  }, numeric(1))
  list(stat = obs, boot = stars, crit95 = unname(quantile(stars, 0.95)),
       p_value = (1 + sum(stars >= obs)) / (1 + B), scheme = scheme, B = B)
}
```

The $(1 + \#)/(1 + B)$ form is exactly valid in finite samples. Choose $B$ so
that $\alpha(1+B)$ is an integer: 499, 999, 1999. Not 500.

```{r boot-run}
bt_wild <- boot_supLR(r, best$p, best$d, scheme = "wild",     mlag = MLAG)
bt_resi <- boot_supLR(r, best$p, best$d, scheme = "residual", mlag = MLAG)
data.frame(
  scheme  = c("wild", "residual", "naive chi^2"),
  crit95  = round(c(bt_wild$crit95, bt_resi$crit95, qchisq(0.95, best$p + 1)), 3),
  p_value = round(c(bt_wild$p_value, bt_resi$p_value,
                    pchisq(bt_wild$stat, best$p + 1, lower.tail = FALSE)), 4))
```

```{r boot-plot, fig.height=3.6}
hist(bt_wild$boot, breaks = 40, freq = FALSE, col = "grey85", border = "white",
     xlab = "sup-LR", main = "Bootstrap null vs the naive chi-square")
curve(dchisq(x, best$p + 1), add = TRUE, col = "red", lwd = 2)
abline(v = bt_wild$stat, lty = 2, lwd = 2)
legend("topright", c("bootstrap null","chi-square (naive)","observed"),
       col = c("grey60","red","black"), lwd = c(6,2,2), lty = c(1,1,2), bty = "n")
```

**Report the $p$-value together with the null DGP that generated it.** A
threshold test is only as credible as its null. If the wild-bootstrap $p$-value
is large while the residual-bootstrap $p$-value is tiny, you have found GARCH.

### What happens if you cheat

Fixing $c = \hat c$ inside the loop removes the supremum. The bootstrapped
statistic becomes a pointwise LR, which is asymptotically $\chi^2$. Since
$\sup_c \mathrm{LR}^*(c) \ge \mathrm{LR}^*(\hat c)$ always, the simulated null is
stochastically too small and the $p$-value too small.

```{r cheat}
cheat <- vapply(seq_len(200), function(b) {
  ys <- sim_ar_null(fit_ar(r, best$p, mlag = MLAG), n, scheme = "wild")
  ar_s <- fit_ar(ys, best$p, mlag = MLAG)
  s1 <- setar_ssr(ys, best$p, best$d, best$c_hat, mlag = MLAG)  # c FIXED
  ar_s$n * (log(ar_s$ssr) - log(s1))
}, numeric(1))
c(correct_95  = round(bt_wild$crit95, 3),
  cheating_95 = round(unname(quantile(cheat, 0.95, na.rm = TRUE)), 3),
  chisq_95    = round(qchisq(0.95, best$p + 1), 3))
```

The middle number sits close to the $\chi^2$ value and far below the correct
one. That gap *is* the size distortion.

---

# Forecasting

## One step: easy. $h$ steps: not.

For a SETAR at $h=1$ the regime is known --- you observe $y_{T+1-d}$. At $h=2$
the regime depends on the *random* $y_{T+2-d}$, and since the map $f$ is
nonlinear, $\mathbb{E}[f(Y)] \neq f(\mathbb{E}[Y])$. Iterating the point forecast
through the skeleton is simply wrong, and it discards the interesting object: the
predictive density, which is a mixture and is often skewed.

```{r setar-fc}
setar_forecast1 <- function(fit, ylast) {
  b <- if (ylast[fit$d] <= fit$c) fit$coef1 else fit$coef2
  as.numeric(b[1] + sum(b[-1] * ylast[1:fit$p]))
}

setar_forecast_mc <- function(fit, yhist, h, M = 5000, resid_pool = NULL) {
  pmax <- max(fit$p, fit$d); paths <- matrix(NA_real_, M, h)
  for (m in 1:M) {
    yl <- rev(tail(yhist, pmax))                 # yl[1] = y_T
    for (j in 1:h) {
      low <- yl[fit$d] <= fit$c
      b  <- if (low) fit$coef1 else fit$coef2
      sg <- if (low) fit$sigma1 else fit$sigma2
      e  <- if (is.null(resid_pool)) rnorm(1) else sample(resid_pool, 1)
      yn <- b[1] + sum(b[-1] * yl[1:fit$p]) + sg * e
      yl <- c(yn, yl[-pmax]); paths[m, j] <- yn
    }
  }
  paths
}
```

## Markov switching: closed form for the regime, mixture for the density

$$
\xi_{T+h|T} = (P')^h \xi_{T|T}, \qquad
\mathrm{Var}(r_{T+h}\mid\mathcal{F}_T)
= \underbrace{\sum_j \xi_j \sigma_j^2}_{\text{within}}
+ \underbrace{\sum_j \xi_j (m_j - \bar m)^2}_{\text{across}}.
$$

```{r ms-fc}
ms_forecast1 <- function(th, xi_T, ylast) {
  P <- matrix(c(th$p11, 1 - th$p11, 1 - th$p22, th$p22), 2, 2, byrow = TRUE)
  xi1 <- as.vector(crossprod(P, xi_T))
  m_j <- th$mu + th$phi * ylast; mbar <- sum(xi1 * m_j)
  list(xi = xi1, mean = mbar,
       var = sum(xi1 * th$sig^2) + sum(xi1 * (m_j - mbar)^2),
       comp_mean = m_j, comp_sd = th$sig)
}
ms_dens <- function(x, fc) sum(fc$xi * dnorm(x, fc$comp_mean, fc$comp_sd))
ms_cdf  <- function(x, fc) sum(fc$xi * pnorm(x, fc$comp_mean, fc$comp_sd))
ms_quantile <- function(a, fc)
  uniroot(function(x) ms_cdf(x, fc) - a,
          c(min(fc$comp_mean) - 12*max(fc$comp_sd),
            max(fc$comp_mean) + 12*max(fc$comp_sd)))$root

fc <- ms_forecast1(th, flt$xi_filt[n, ], r[n])
c(xi1 = fc$xi[1], mean = fc$mean, sd = sqrt(fc$var),
  VaR1_mixture = ms_quantile(0.01, fc),
  VaR1_gaussian = qnorm(0.01, fc$mean, sqrt(fc$var)))
```

The Gaussian comparator has, by construction, the **same mean and the same
variance**. It therefore has identical MSFE and an identical Mincer--Zarnowitz
regression. Its 1% VaR is different. Point-forecast evaluation is structurally
blind to this; that is the argument for evaluating densities.

```{r dens-plot, fig.height=3.4}
xs <- seq(fc$mean - 6*sqrt(fc$var), fc$mean + 6*sqrt(fc$var), length.out = 600)
plot(xs, sapply(xs, ms_dens, fc = fc), type = "l", lwd = 2, col = "firebrick",
     xlab = "r_{T+1} (%)", ylab = "density",
     main = "Predictive density: MS mixture vs matched Gaussian")
lines(xs, dnorm(xs, fc$mean, sqrt(fc$var)), lwd = 2, col = "steelblue")
legend("topright", c("MS mixture","matched Gaussian"),
       col = c("firebrick","steelblue"), lwd = 2, bty = "n")
```

---

# Rolling out-of-sample exercise

Three design rules, each easy to break and expensive to break:

* Re-estimate **everything** inside the window --- $p$, $d$, $c$, and the MS
  parameters. Choosing $p$ once on the full sample is look-ahead bias.
* Nonlinear models need **long windows**. With a 500-day window the crash regime
  may never appear in-sample, and the model degenerates to its linear special
  case.
* `refit_every` re-estimates only every $k$ origins. This mimics a real desk and
  keeps run time finite. Set it to 1 for the textbook-pure version.

```{r oos, cache=TRUE}
R_win <- CFG$window
origins <- seq(R_win, n - 1)
if (length(origins) > CFG$max_origins) origins <- tail(origins, CFG$max_origins)
P_oos <- length(origins); actual <- r[origins + 1]
f_ar <- f_st <- f_ms <- rep(NA_real_, P_oos)
ls_ar <- ls_st <- ls_ms <- pit <- rep(NA_real_, P_oos)
par_ms <- fit_ms$par

for (i in seq_along(origins)) {
  t0 <- origins[i]; yw <- r[(t0 - R_win + 1):t0]
  if ((i - 1) %% CFG$refit_every == 0) {
    arw <- fit_ar(yw, p_hat, mlag = MLAG)
    gw  <- setar_grid(yw, p_hat, best$d, mlag = MLAG)
    stw <- setar_fit(yw, p_hat, best$d, gw$c_hat, mlag = MLAG)
    msw <- try(optim(par_ms, ms_negll, y = yw, method = "BFGS",
                     control = list(maxit = 1000)), silent = TRUE)
    if (!inherits(msw, "try-error")) par_ms <- msw$par
    thw <- ms_unpack(par_ms)
    if (thw$sig[1] > thw$sig[2]) {
      par_ms[c(1,2)] <- par_ms[c(2,1)]; par_ms[c(4,5)] <- par_ms[c(5,4)]
      par_ms[c(6,7)] <- par_ms[c(7,6)]; thw <- ms_unpack(par_ms)
    }
  }
  fw <- ms_filter(par_ms, yw); if (is.null(fw)) next
  yl <- rev(tail(yw, max(p_hat, best$d)))
  f_ar[i] <- arw$coef[1] + sum(arw$coef[-1] * yl[1:p_hat])
  f_st[i] <- setar_forecast1(stw, yl)
  sg_st   <- if (yl[stw$d] <= stw$c) stw$sigma1 else stw$sigma2
  fci     <- ms_forecast1(thw, fw$xi_filt[length(yw), ], yw[length(yw)])
  f_ms[i] <- fci$mean
  ls_ar[i] <- dnorm(actual[i], f_ar[i], sqrt(arw$sigma2), log = TRUE)
  ls_st[i] <- dnorm(actual[i], f_st[i], sg_st, log = TRUE)
  ls_ms[i] <- log(ms_dens(actual[i], fci))
  pit[i]   <- ms_cdf(actual[i], fci)
}

ok <- complete.cases(f_ar, f_st, f_ms)
e_ar <- actual[ok] - f_ar[ok]; e_st <- actual[ok] - f_st[ok]
e_ms <- actual[ok] - f_ms[ok]
data.frame(model = c("AR","SETAR","MS"),
           MSFE = round(c(mean(e_ar^2), mean(e_st^2), mean(e_ms^2)), 5),
           MAE  = round(c(mean(abs(e_ar)), mean(abs(e_st)), mean(abs(e_ms))), 5),
           logscore = round(c(mean(ls_ar[ok]), mean(ls_st[ok]),
                              mean(ls_ms[ok])), 4))
```

Expected pattern: MSFE essentially tied; log score clearly better for the MS
model. **Nonlinear models earn their keep in the density and in the signs, not in
the mean squared error.**

---

# Forecast evaluation

## Diebold--Mariano --- and when it is illegal

```{r dm}
dm_test <- function(e1, e2, h = 1, power = 2, hln = TRUE) {
  d <- abs(e1)^power - abs(e2)^power; P <- length(d); dbar <- mean(d)
  g <- acf(d, lag.max = max(h - 1, 1), type = "covariance", plot = FALSE)$acf
  lrv <- g[1] + if (h > 1) 2 * sum(g[2:h]) else 0
  stat <- dbar / sqrt(lrv / P)
  if (hln) {
    stat <- stat * sqrt((P + 1 - 2*h + h*(h - 1)/P) / P)
    pv <- 2 * pt(-abs(stat), df = P - 1)
  } else pv <- 2 * pnorm(-abs(stat))
  c(DM = stat, p = pv, dbar = dbar)
}
```

An AR($p$) is **nested** in a SETAR: set $\phi^{(1)} = \phi^{(2)}$. Under the
null the two forecasts converge to each other, so $\bar d \to 0$ *and*
$\mathrm{Var}(d_t) \to 0$ at the same rate. The DM statistic is degenerate and
badly undersized: "fail to reject" carries almost no information. Worse, under
the null the larger model's forecasts contain pure estimation noise, which
mechanically raises its MSFE and pushes $\bar d$ the wrong way.

Clark--West subtracts exactly that noise term, $(\hat r_1 - \hat r_2)^2$. It is a
**one-sided** test.

```{r cw}
cw_test <- function(y, f_small, f_large, h = 1) {
  ft <- (y - f_small)^2 - ((y - f_large)^2 - (f_small - f_large)^2)
  P <- length(ft); fbar <- mean(ft)
  g <- acf(ft, lag.max = max(h - 1, 1), type = "covariance", plot = FALSE)$acf
  lrv <- g[1] + if (h > 1) 2 * sum(g[2:h]) else 0
  stat <- fbar / sqrt(lrv / P)
  c(CW = stat, p_one_sided = pnorm(stat, lower.tail = FALSE), fbar = fbar)
}

rbind(
  "DM  SETAR vs MS (non-nested: legitimate)" = round(dm_test(e_st, e_ms), 4),
  "DM  AR vs SETAR (NESTED: uninformative)"  = round(dm_test(e_ar, e_st), 4))
round(cw_test(actual[ok], f_ar[ok], f_st[ok]), 4)
```

## Directional accuracy: Pesaran--Timmermann

For returns, sign accuracy often matters more than MSFE. A model can be poor at
levels and profitable at signs.

```{r pt}
pt_test <- function(y, f) {
  keep <- y != 0 & f != 0; y <- y[keep]; f <- f[keep]; P <- length(y)
  Phat <- mean(sign(y) == sign(f))
  py <- mean(y > 0); pf <- mean(f > 0)
  Ps <- py*pf + (1 - py)*(1 - pf)
  vP  <- Ps*(1 - Ps)/P
  vPs <- ((2*py - 1)^2 * pf*(1 - pf) + (2*pf - 1)^2 * py*(1 - py) +
            4*py*pf*(1 - py)*(1 - pf)/P) / P
  stat <- (Phat - Ps) / sqrt(max(vP - vPs, .Machine$double.eps))
  c(hit_rate = Phat, PT = stat, p = pnorm(stat, lower.tail = FALSE))
}
rbind(AR = round(pt_test(actual[ok], f_ar[ok]), 4),
      SETAR = round(pt_test(actual[ok], f_st[ok]), 4),
      MS = round(pt_test(actual[ok], f_ms[ok]), 4))
```

## Density evaluation: the PIT

If the density forecast is correctly specified, $u_t = F_t(r_t)$ is i.i.d.
$U(0,1)$. A **U-shaped** histogram means the intervals are too narrow (tails
underestimated); a **hump** means they are too wide. Uniformity and independence
are separate hypotheses and both must hold.

```{r pit, fig.height=3.4}
u <- pit[ok]
hist(u, breaks = 20, freq = FALSE, col = "grey85", border = "white",
     xlab = "PIT", main = "PIT of the MS density forecast")
abline(h = 1, col = "red", lwd = 2)
ks.test(u, "punif")
Box.test(u - mean(u), lag = 10, type = "Ljung-Box")
```

---

# Checklist before you believe any of this

1. Was the AR order re-selected inside the rolling window? *(look-ahead bias)*
2. Did the bootstrap re-run the grid search over $c$ in **every** draw?
3. Is the bootstrap null DGP reported alongside the $p$-value?
4. Is the DM test being applied to nested models? *(it must not be)*
5. Is a nonlinear model being judged on MSFE alone? *(it should not be)*
6. Does the MS crisis dating line up with 2008, 2020 and 2022?
7. Is the profile-SSR minimum sharp, or is $\hat c$ barely identified?

```{r session}
sessionInfo()
```
